Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-05-22 manuscript/311 manuscript.tex
Daily updates

\usepackage[margin = 1in]{geometry}
\usepackage{amsmath}
\allowdisplaybreaks
\usepackage{authblk}
\usepackage{setspace}
\usepackage{natbib}
\usepackage{hyperref}\hypersetup{colorlinks=true, citecolor=blue}
\usepackage{parskip}
\usepackage{textgreek}
\usepackage{graphicx}
\usepackage{booktabs} % For better looking tables
\usepackage{siunitx} % For alignment of numbers
\usepackage{placeins}
\usepackage{caption}  % For custom captions

@@

The daily tend over that same 10-year period is also interesting, as it shows both significant spikes
and significant dips; a noisy graph. While the spikes in 2020 and 2021 appear COVID related, we are unable to 
ascertain any reason for the significant spike on Saturday, 2023-09-02 where 115,205 SRs were created.

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.3]{Daily.png}
  \caption{SR counts by day for a 10-year period}
  \label{fig:daily-counts}
\end{figure}

@@

Of note, the lowest SR counts-per-day occur on Christmas  period (+/- a day), in 5 of the Bottom 10 SR counts-per-day. The summer moths and early
fall months make up all of the Top 10 SR counts-per-day. As previously noted, the all-time daily high for SRs occurred on Tuesday, 2020-08-04 with the 10-year
low number occurs on Christmas Day in 2016, a Sunday.

@@

@@

	\subsubsection{Fields not in the Data Dictionary}
	The 311 SR \href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary} identifies 41 data columns (fields) along with some
	related information for each column. However, when you either download the data or explore it via the online portal,
	you will immediately note that there are 47 fields. The additional six fields are not addressed in the Data Dictionary, and show up on the Column Manager
	page as ``@computed\_region\_xxxx\_xxxx''. To some extent one can infer what the these @computed\_fields are, but it is not possible to know
	for certain. Here is a screenshot of the NYC Open Data Portal ``Column Manager'' page showing those columns:

@@

	The following fields were checked for their specified data types:
		\begin{itemize}
		\item created\_date, closed\_date, due\_date, and resolution\_action\_updated\_date are all in proper date format.
		\item zip\_codes and incident\_zip: All are 5 numeric digits except for 2 non-numeric entries for incident\_zip (``na'' and ``N/A'').
		\item x\_coordinate\_state\_plane \&  y\_coordinate\_state\_plane (a NY State geo-location system) are all numeric.
		\item latitude \& longitude are both numeric.
		\item community\_districts, borough\_boundaries, city\_council\_district, police\_precinct, and police\_precincts are all numeric.
		\item Free-form text fields that contain a ``,'' (e.g. incident\_address, resolution\_description) are enclosed in quotation marks.
		\end{itemize}

@@

\subsection{Data type correctness}

\subsection{Invalid values}

>>>>

\usepackage[margin = 1in]{geometry}
\usepackage{amsmath}
\allowdisplaybreaks
\usepackage{authblk}
\usepackage{setspace}
\usepackage{natbib}
\usepackage{hyperref}\hypersetup{colorlinks=true, citecolor=blue}
\usepackage{parskip}
\usepackage{textgreek}
\usepackage{graphicx}
\usepackage{booktabs} % For better looking tables
\usepackage{siunitx} % For alignment of numbers
\usepackage{placeins}
\usepackage{caption}  % For custom captions
\usepackage{float}   % Required for the H specifier

@@

The daily tend over that same 10-year period is also interesting, as it shows both significant spikes
and significant dips; a noisy graph. While the spikes on 2020-08-04 (24,415 SRs) and 2020-08-05 (19,560 SRs) are COVID related, we are unable to 
ascertain any reason for the significant spike on Saturday, 2023-09-29 where 17,962 SRs were created.

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.35]{Daily.png}
  \caption{SR counts by day for a 10-year period}
  \label{fig:daily-counts}
\end{figure}

@@

Of note, the lowest SR counts-per-day occur on Christmas  period (+/- a day), in 5 of the Bottom 10 SR counts-per-day. The summer moths and early
fall months make up all of the Top 10 SR counts-per-day. As previously noted, the all-time daily high for SRs occurred on Tuesday, 2020-08-04 with the 10-year
low number occurs on Christmas Day in 2016, a Sunday. Note also that the difference between the peak day (24,415) and the minimum day (2965) varies
by a factor of 8X, and the aveage daily count of 7489 -- all of which creates a challenging  system capacity design issue.

@@

What are the characteristics of the 311 SR data set. Here are some insights:

\begin{itemize}
	\item 47 columns of data for each row, exportable as a CSV file
	\item There are four date fields (created, closed, updated, due)
	\item There are three borough fields; two of which we believe to be duplicates
	\item Two zip code fields, but not duplicates
	\item Seven street fields; one pair of which we believe to be duplicates
	\item Agency name and Agency abbreviation
	\item Two Police Precinct fields; not duplicates
	\item In addition to street/city, there are three other location fields (lat/long), NY State plane, and Block \#
	\item Incident address (25 Grymes Hill Road) and Street Name (Grymes Hill Road)
	\item One form tex fielt; resolution description, which supports 934 characters of input
\end{itemize}

@@

	\subsubsection{Fields not in the Data Dictionary}
	The 311 SR \href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary}
	 identifies 41 data columns (fields) along with some
	related information for each column. However, when you either download the data or explore it via the online portal,
	you will immediately note that there are 47 fields. The additional six fields are not addressed in the Data Dictionary, and show up on the Column Manager
	page as ``@computed\_region\_xxxx\_xxxx''. To some extent one can infer what the these @computed\_fields are, but it is not possible to know
	for certain. These fields are:
	\begin{itemize}
		\item zip\_codes
		\item community\_districts
		\item borough\_boundaries
		\item city\_council\_districts
		\item police\_precincts
		\item police\_precinct 
	\end{itemize}

	Here is a screenshot of the NYC Open Data Portal ``Column Manager'' page showing those columns:

@@

	The following fields were checked for their specified data types:

		\begin{itemize}
		\item created\_date, closed\_date, due\_date, and resolution\_action\_updated\_date are all in proper date format.
		\item zip\_codes and incident\_zip: All are 5 numeric digits except for 2 non-numeric entries for incident\_zip (``na'' and ``N/A'').
		\item x\_coordinate\_state\_plane \&  y\_coordinate\_state\_plane (a NY State geo-location system) are all numeric.
		\item latitude \& longitude are both numeric.
		\item community\_districts, borough\_boundaries, city\_council\_district, police\_precinct, and police\_precincts are all numeric.
		\item Free-form text fields that contain a ``,'' (e.g. incident\_address, resolution\_description) are enclosed in quotation marks.
		\end{itemize}

@@

Understanding the absence of data in each fields is an important factor when undertaking analysis. For example, if you wanted to see if the SRs were
closed before or after their due\_date, you would be challenged as 99.57\% of the due\_date field is blank. (Although with such a large dataset
that still leaves 24,411 values, all of which are N/A, which calls into question are these missing values, or is the specification of a
due\_date simply not applicable for the vast majority of SRs?)

When counting the various fields for blank or N/A values, the dataset appears to divide into three groups: Mostly Empty, Partially Empty, and Few/None Empty.
The Mostly Empty category ranges from 93-99.9\% blank and includes such fields as taxi\_company due\_date, pickup\_location, and landmark. The Partially 
Empty includes such fields as location\_type, borough, and cross\_street. And the Few/None Empty includes created\_date, complaint\_type, agency, and status.
In some cases it may make sense to inquiry as to why some fields are frequently blank. Is the data difficult to capture? Does it pertain to only a small set
of complaint\_type or Agencies? And if the data is truly that sparse, does it make sense to even collect it. For example, you would think that every complaint
that every SR that has a geographical component, such as an address, would also have a corresponding police\_precinct and community\_district since these
are all-inclusive. That it would have a lat/long.  But indeed these fields do have small, but nonetheless significant empty entries.

During the analysis, it was determined that the following fields are mandatory, and any rows missing these fields is removed from the dataset (which very, very
rarely occurred fortunately): created\_date, agency, complaint\_type, and unique\_key.

\begin{table}[ht]
\centering
\caption{Blank and N/A entries by field}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lrrrl}
	\toprule
	\textbf{Field} & \textbf{Total Empty} & \textbf{Percent Empty} & \textbf{NA Count} & \textbf{Blanks Only} \\
	\midrule
		taxi\_company\_borough & 6391341 & 99.94 & 0 & 6391341 \\
		road\_ramp & 6379080 & 99.75 & 0 & 6379080 \\
		vehicle\_type & 6376266 & 99.71 & 0 & 6376266 \\
		due\_date & 6370497 & 99.62 & 6370497 & 0 \\
		bridge\_highway\_direction & 6367626 & 99.57 & 0 & 6367626 \\
		bridge\_highway\_name & 6340197 & 99.14 & 0 & 6340197 \\
		bridge\_highway\_segment & 6340190 & 99.14 & 0 & 6340190 \\
		taxi\_pick\_up\_location & 6329954 & 98.98 & 0 & 6329954 \\
		facility\_type & 5966588 & 93.30 & 0 & 5966588 \\
		landmark & 2635736 & 41.22 & 0 & 2635736 \\
		intersection\_street\_1 & 2143847 & 33.52 & 0 & 2143847 \\
		intersection\_street\_2 & 2140707 & 33.48 & 0 & 2140707 \\
		cross\_street\_1 & 1846594 & 28.88 & 0 & 1846594 \\
		cross\_street\_2 & 1846008 & 28.87 & 0 & 1846008 \\
		location\_type & 798775 & 12.49 & 0 & 798775 \\
		bbl & 768109 & 12.01 & 0 & 768109 \\
		city & 332086 & 5.19 & 0 & 332086 \\
		street\_name & 280463 & 4.39 & 0 & 280463 \\
		incident\_address & 280268 & 4.38 & 0 & 280268 \\
		closed\_date & 248839 & 3.89 & 248839 & 0 \\
		resolution\_description & 132333 & 2.07 & 0 & 132333 \\
		zip\_codes & 123167 & 1.93 & 0 & 123167 \\
		borough\_boundaries & 101453 & 1.59 & 0 & 101453 \\
		community\_districts & 101445 & 1.59 & 0 & 101445 \\
		city\_council\_districts & 101445 & 1.59 & 0 & 101445 \\
		police\_precincts & 101445 & 1.59 & 0 & 101445 \\
		police\_precinct & 101416 & 1.59 & 0 & 101416 \\
		latitude & 99778 & 1.56 & 0 & 99778 \\
		longitude & 99778 & 1.56 & 0 & 99778 \\
		location & 99778 & 1.56 & 0 & 99778 \\
		x\_coordinate\_state\_plane & 99668 & 1.56 & 0 & 99668 \\
		y\_coordinate\_state\_plane & 98822 & 1.55 & 0 & 98822 \\
		resolution\_action\_updated\_date & 91399 & 1.43 & 91399 & 0 \\
		incident\_zip & 83220 & 1.30 & 2 & 83218 \\
		descriptor & 74645 & 1.17 & 0 & 74645 \\
		address\_type & 38655 & 0.60 & 0 & 38655 \\
		unique\_key & 0 & 0.00 & 0 & 0 \\
		agency & 0 & 0.00 & 0 & 0 \\
		agency\_name & 0 & 0.00 & 0 & 0 \\
		complaint\_type & 0 & 0.00 & 0 & 0 \\
		status & 0 & 0.00 & 0 & 0 \\
		community\_board & 0 & 0.00 & 0 & 0 \\
		borough & 0 & 0.00 & 0 & 0 \\
		open\_data\_channel\_type & 0 & 0.00 & 0 & 0 \\
		park\_facility\_name & 0 & 0.00 & 0 & 0 \\
		park\_borough & 0 & 0.00 & 0 & 0 \\
		created\_date & 0 & 0.00 & 0 & 0 \\
	\bottomrule
\end{tabular}
}
\end{table}

Here is a graphic depiction of total empty (blank + N/As) for each fields. You can see the natural grouping into the Most, Some, and Few categories.

\begin{figure}[htbp]
  \centering
	  \includegraphics[scale=0.7]{BlankFields.png}
	  \caption{Number and Percentage of Empty Entries}
	  \label{fig:blank_fields}
\end{figure}

\subsection{Data type correctness}
All data fields were subjected to a test of ``correctness'' according to their ``type'' as specified
in the Data Dictionary.  Of the nearly 300 million data fields to check (6.5 million rows, 47 fields per row), there were only two  values that failed to test for the
proper data type; those being two text entries in the incident\_zip filed (``na'', and N/A). All date fields were dates. All number fields were numeric. All text
fields were character.

\subsection{Checking for Valid Values}
Testing for valid and invalid values revealed several fairly serious issues associated with select fields. Accordingly, an analyst would need to be
very diligent if using these fields for analysis. In all likelihood, the rows with the invalid field values would likely need to be removed from the
analysis.

There were a variety of tests for validity that were applied individually to specific fields. Here are the ``good'' results:

\begin{itemize}
		\item Latitude and Longitude fields were tested to ensure all fell within the geographic boundaries of the City of New York. All were.
		\item The unique\_key field was in fact unique.
		\item Many fields have a domain of acceptable values. Often these values are determined from common usage or by examining
		larger datasets. Unfortunately, these domains of acceptable values are not specified in the Data Dictionary. These fields all tested as
		compliant with their domain of acceptable values:
			\begin{itemize}
				\item address\_type
				\item status
				\item borough, borough\_boundaries, \& park\_borough 
				\item data\_channel
				\item vehicle\_type
				\item city\_council\_district
			\end{itemize}
\end{itemize}

 Unfortunately some fields did prove to be quite problematic when comparing the values to the domain of legal values.
 For example, all zip codes (two fields: zip\_codes and incident\_zip) should be valid as defined by the USPS database
 which contains over 44,000 valid zip codes.

 The computed field zip\_codes proved especially problematic with 58\% (3.6 million) of the field entries being invalid.
This high number indicates that the computation of this computed fields is highly inaccurate. We would recommend
dropping this field for future analytic efforts.

Furthermore, the field incident\_zip while having only .07\% invalid entries, that is still 4163 such errors. As zip code is
a primary measure of many NYC city services, and with a freely available database for which to validate entries
against, it seems an oversight that any such errors could creep into the system. Some invalid entries are 
clearly incorrect just by observation, e.g. 10000, 12345, 11111, etc.

Because zip codes play a significant role in analyzing City services, these errors can be quite serious. It's useful in 
such cases to observe the breakdown of these errors by the responsible Agency.

The breakdown of the invalid entries in the zip\_code, sorted by Agency shows that the distribution by percentage
almost precisely mirrors the overall breakdown of SRs by Agency. This indicates a systemic problem, and since the
zip\_code fields is one of the computed fields, it appears these errors are caused by incorrect computations rather
than by any Agency mistake.

The  errors in the incident\_zip field are more troubling, even though they are, percentage-wise, small. Here is a graphic 
illustrating how these errors occur by Agency. Here we can see that the majority of these errors lie in the Dept of Transportation,
NYPD, Taxi \& Limo Commission (TLC), and the Economic Development Council (EDC) among others. This representation
does not follow the full SR distribution indicating that these errors are likely generated by an incorrect
process or application residing at those particular Agencies; it is not a systemic problem.

\begin{figure}[htbp]
  \centering
	  \includegraphics[scale=0.7]{invalid_incident_zip.png}
	  \caption{Invalid incident-zip by Agency}
	  \label{fig:invalid_incident_zip}
\end{figure}

A curious case also exists when examining the two nearly identical fields - police\_precincts and police\_precinct.
